Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems

### 2017-12-05 dissipative_hamiltonian_revision/chapters/3.model.tex
dissipative Hmiltonian revision submitted

Suppose that a reduced symplectic basis $A \in \mathbb R^{2n\times 2k}$ is provided with $k \ll n$. The approximate solution to (\ref{eq:2.1}) in this basis is expressed as
\begin{equation} \label{eq:3.1}
	z =Ay,
\end{equation}
where $y\in \mathbb R^{2k}$ are the coordinates of the approximation with respect to the basis $A$. Substituting (\ref{eq:3.1}) into (\ref{eq:2.1}) yields
\begin{equation} \label{eq:3.2}
	A \dot y = \mathbb{J}_{2n} \nabla_{z} H(A y).
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we obtain
\begin{equation} \label{eq:3.3}
	\dot y = A^+ \mathbb{J}_{2n} (A^+)^T \nabla_{y} H(A y).
\end{equation}
As $(A^+)^T$ is a symplectic matrix it implies that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, as $\tilde H (y) = H(Ay)$, we recover the reduced system
\begin{equation} \label{eq:3.3.1}
	\begin{aligned}
	\dot {y}(t) &= \mathbb J_{2k} \nabla_{y} \tilde H, \\
	y(0) &= A^+ z_0,
	\end{aligned}
\end{equation}
Equation (\ref{eq:3.3.1}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}). Conventional model reduction techniques, e.g. Galerkin or Petrov-Galerkin methods \cite{hesthaven2015certified,quarteroni2015reduced}, do not yield a Hamiltonian reduced system and the reduced system does not necessarily preserve the conservation law in Theorem \ref{theorem:2.1} which results in a qualitatively wrong and often unstable solution \cite{Peng:2014di}. On the other hand the reduced system, obtained by the symplectic Galerkin projection, is a Hamiltonian system, and the system energy is therefore preserved over time \cite{Peng:2014di}. The following theorem guarantees the boundedness of the solution of the reduced system obtained by the symplectic Galerkin projection for quadratic Hamiltonians.

@@

\subsection{The Greedy Approach to Symplectic Basis Generation} \label{sec:3.2}
Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. Suppose that at the $k$-th generic step of the greedy basis selection, an ortho-symplectic basis $A_{2k} = \{ e_1,\dots,e_k,f_1,\dots,f_k\}$ of size $2k$ is provided. The first step in an iteration of the greedy basis selection comprises finding $t^{k+1}\in[0,T]$ such $z(t^{k+1})$ is worst approximated by the current reduced space. In other words
\begin{equation} \label{eq:3.4}
	t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|.
\end{equation}
Suppose that $e_{k+1}$ is the vector obtained by $\mathbb J_{2n}$-orthogonalization (the symplectic Gram-Schmidt process \cite{Salam2014}) of $z(t^{k+1})$ with respect to $A_{2k}$. Then the enriched basis $A_{2k+2}$ takes the form
\begin{equation}
	A_{2k+2} = \{ e_1,\dots,e_k, e_{k+1},f_1,\dots,f_k,\mathbb J_{2n}^T e_{k+1}\}.
\end{equation}
It is easily checked that $A_{2k+2}$ is ortho-symplectic. Furthermore, it is shown in \cite{Maboudi:2016} that under natural assumptions on the solution manifold $\mathcal M$, the greedy method converges exponentially fast.

@@

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:added1}
{\bf Input:} Tolerated projection error $\delta$, initial condition $ z_0$
\begin{enumerate}
\item $t^1 \leftarrow t=0$
\item $e_1 \leftarrow z_0$
\item $A \leftarrow [e_1,\mathbb J^T_{2n}e_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\| z(t) - A_{2k}{A_{2k}}^+z(t) \| > \delta$ for all $t \in [0,T]$
\item \hspace{0.5cm} $t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|$
\item \hspace{0.5cm} $\mathbb J_{2n}$-orthogonalize $ z(t^{k+1})$ to obtain $e_{k+1}$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

@@

Consider a dissipative Hamiltonian system of the form (\ref{eq:2.4}) with a quadratic Hamiltonian, $H(z) = z^TK^TKz$. Since $K^TK$ is symmetric and positive definite, it has a unique Cholesky factorization $K^TK = L^T L$ where $L$ is upper triangular \cite{strang09}. So we can write 
\begin{equation} \label{eq:3.7}
H(z) = z^T L^T L z.
\end{equation}
Further, suppose that the solution $z(t)$ lies on a low-dimensional symplectic subspace such that $z = Ay$, where $A$ is an ortho-symplectic matrix of the size $2n\times 2k$ and $y$ is the expansion coefficients of $z$ in the basis of $A$. The Hamiltonian $H$ then takes the form
\begin{equation} \label{eq:3.8}
	H(z) = H(Ay) = y^T A^T L^T L A y.
\end{equation}
Since $A^T L^T L A$ is symmetric and positive definite, it too has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is an upper triangular matrix of size $2k \times 2k$. Writing (\ref{eq:2.4}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
\begin{equation} \label{eq:3.9}
		A\dot{y}(t) = \mathbb J_{2n} L^T f(t),
\end{equation}
together with the complementary equation
\begin{equation} \label{eq:3.10}
	f(t) + \sqrt 2 \cdot \sqrt{\chi} \phi(t,0) = LAy.
\end{equation}
Multiplying (\ref{eq:3.9}) with $A^+$ and (\ref{eq:3.10}) with $A^T$ yields
\begin{align} \label{eq:3.11}
	& \dot y(t) = \mathbb J_{2k} A^T L^T f(t), \\
	& A^T f(t) + \sqrt{2} A^T \sqrt{\chi} \phi(t,0) = A^T L A y,
\end{align}
where we use the fact that $A^+\mathbb J_{2n} = \mathbb{J}_{2k} A^T$. If we define $f = A \tilde f$, $\phi = A \tilde \phi$, $\theta = A\tilde \theta$ and the \emph{reduced susceptibility} as $\tilde \chi = A^T \chi A$ we recover the reduced Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:3.12.a} & \dot{y}(t) = \mathbb J_{2k} {\tilde L}^T \tilde f(t), \\
		\label{eq:3.12.c} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),\\
		\label{eq:3.12.b} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t),
\end{align}
\end{subequations}
together with the auxiliary equation
\begin{equation} \label{eq:3.13}
	\tilde f(t) + \sqrt{2} \sqrt{\tilde \chi} \tilde \phi(t,0) = \tilde L y.
\end{equation}
Equations (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
\begin{equation}
	\dot {y} = \mathbb J_{2k} \tilde L^T \tilde f(t), \quad \tilde f(t) + \int_0^t \tilde \chi\cdot \tilde f(s)\ ds = \tilde L y.
\end{equation}
Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a})-(\ref{eq:3.12.b}).

We point out that the transformation that connects (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is given by
\begin{equation}
	\mathcal A = \begin{pmatrix}
		A& 0 \\
		0& A
	\end{pmatrix} : \mathbb R^{2n} \times \mathcal H^{2n} \to \mathbb R^{2k} \times \mathcal H^{2k}.
\end{equation}
This is a symplectic transformation, since $\mathcal A^T \mathcal J_{2n} \mathcal A = \mathcal J_{2k}$. Furthermore, the dissipation of energy in the reduced system only depends on the reduced susceptibility. Thus, the choice of $A$ should be independent of the hidden strings $(\phi, \theta)$. In other words, if the reduced space is chosen to be a symplectic subspace, then the actions of model reduction and Hamiltonian extension commute. We summarize the algorithm for model reduction of dissipative Hamiltonian systems in Algorithm \ref{alg:3.1}.

\begin{algorithm}
\caption{The Reduced Dissipative Hamiltonian Method (RDH)} \label{alg:3.1}
\begin{enumerate}
	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to the original TDD system (\ref{eq:2.4}).
	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
	\item Construct an ortho-symplectic basis $A$.
	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.b})
\end{enumerate}
\end{algorithm}

@@

>>>>

Suppose that a reduced symplectic basis $A \in \mathbb R^{2n\times 2k}$ is provided with $k \ll n$. The approximate solution to (\ref{eq:2.1}) in this basis is expressed as
\begin{equation} \label{eq:3.1}
	z =Ay,
\end{equation}
where $y\in \mathbb R^{2k}$ are the coordinates of the approximation with respect to the basis $A$. Substituting (\ref{eq:3.1}) into (\ref{eq:2.1}) yields
\begin{equation} \label{eq:3.2}
	A \dot y = \mathbb{J}_{2n} \nabla_{z} H(A y).
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we obtain
\begin{equation} \label{eq:3.3}
	\dot y = A^+ \mathbb{J}_{2n} (A^+)^T \nabla_{y} H(A y).
\end{equation}
As $(A^+)^T$ is a symplectic matrix it implies that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, as $\tilde H (y) = H(Ay)$, we recover the reduced system
\begin{equation} \label{eq:3.3.1}
	\begin{aligned}
	\dot {y}(t) &= \mathbb J_{2k} \nabla_{y} \tilde H, \\
	y(0) &= A^+ z_0.
	\end{aligned}
\end{equation}
Equation (\ref{eq:3.3.1}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}). Conventional model reduction techniques, e.g. Galerkin or Petrov-Galerkin methods \cite{hesthaven2015certified,quarteroni2015reduced}, do not yield a Hamiltonian reduced system and the reduced system does not necessarily preserve the conservation law in Theorem \ref{theorem:2.1} which results in a qualitatively wrong and often unstable solution \cite{Peng:2014di}. On the other hand the reduced system, obtained by the symplectic Galerkin projection, is a Hamiltonian system, and the system energy is therefore preserved over time \cite{Peng:2014di}. The following theorem guarantees the boundedness of the solution of the reduced system obtained by the symplectic Galerkin projection for quadratic Hamiltonians.

@@

\subsection{The Greedy Approach to Symplectic Basis Generation} \label{sec:3.2}
Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. Suppose that at the $k$-th generic step of the greedy basis selection, an ortho-symplectic basis $A_{2k} = \{ e_1,\dots,e_k,f_1,\dots,f_k\}$ of size $2k$ is provided. The first step in an iteration of the greedy basis selection comprises finding $t^{k+1}\in[0,T]$ such that $z(t^{k+1})$ is worst approximated by the current reduced space. In other words
\begin{equation} \label{eq:3.4}
	t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2.
\end{equation}
Suppose that $e_{k+1}$ is the vector obtained by $\mathbb J_{2n}$-orthogonalization (the symplectic Gram-Schmidt process \cite{Salam2014}) of $z(t^{k+1})$ with respect to $A_{2k}$. Then the enriched basis $A_{2k+2}$ takes the form
\begin{equation}
	A_{2k+2} = \{ e_1,\dots,e_k, e_{k+1},f_1,\dots,f_k,\mathbb J_{2n}^T e_{k+1}\}.
\end{equation}
It is easily checked that $A_{2k+2}$ is ortho-symplectic. Furthermore, it is shown in \cite{Maboudi:2016} that under natural assumptions on the solution manifold $\mathcal M$, the greedy method converges exponentially fast.

@@

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:added1}
{\bf Input:} Tolerated projection error $\delta$, initial condition $ z_0$
\begin{enumerate}
\item $t^1 \leftarrow t=0$
\item $e_1 \leftarrow z_0$
\item $A \leftarrow [e_1,\mathbb J^T_{2n}e_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2 > \delta$ for all $t \in [0,T]$
\item \hspace{0.5cm} $t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2$
\item \hspace{0.5cm} $\mathbb J_{2n}$-orthogonalize $ z(t^{k+1})$ to obtain $e_{k+1}$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

@@

Consider a dissipative Hamiltonian system of the form (\ref{eq:2.4}) with a quadratic Hamiltonian, $H(z) = z^TK^TKz$. Since $K^TK$ is symmetric and positive definite, it has a unique Cholesky factorization $K^TK = L^T L$ where $L$ is upper triangular \cite{strang09}. So we can write 
\begin{equation} \label{eq:3.7}
H(z) = z^T L^T L z.
\end{equation}
Further, suppose that the solution $z(t)$ lies on a low-dimensional symplectic subspace such that $z = Ay$, where $A$ is an ortho-symplectic matrix of the size $2n\times 2k$ and $y$ is the expansion coefficients of $z$ in the basis of $A$. The Hamiltonian $H$ then takes the form
\begin{equation} \label{eq:3.8}
	H(z) = H(Ay) = y^T A^T L^T L A y.
\end{equation}
Since $A^T L^T L A$ is symmetric and positive definite, it too has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is an upper triangular matrix of size $2k \times 2k$. Writing (\ref{eq:2.4}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
\begin{equation} \label{eq:3.9}
		A\dot{y}(t) = \mathbb J_{2n} L^T f(t),
\end{equation}
together with the complementary equation
\begin{equation} \label{eq:3.10}
	f(t) + \sqrt 2 \cdot \sqrt{\chi} \phi(t,0) = LAy.
\end{equation}
Multiplying (\ref{eq:3.9}) with $A^+$ and (\ref{eq:3.10}) with $A^T$ yields
\begin{align} \label{eq:3.11}
	& \dot y(t) = \mathbb J_{2k} A^T L^T f(t), \\
	& A^T f(t) + \sqrt{2} A^T \sqrt{\chi} \phi(t,0) = A^T L A y,
\end{align}
where we use the fact that $A^+\mathbb J_{2n} = \mathbb{J}_{2k} A^T$. If we define $f = A \tilde f$, $\phi = A \tilde \phi$, $\theta = A\tilde \theta$ and the \emph{reduced susceptibility} as $\tilde \chi = A^T \chi A$ we recover the reduced Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:3.12.a} & \dot{y}(t) = \mathbb J_{2k} {\tilde L}^T \tilde f(t), \\
		\label{eq:3.12.b} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),\\
		\label{eq:3.12.c} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t),
\end{align}
\end{subequations}
together with the auxiliary equation
\begin{equation} \label{eq:3.13}
	\tilde f(t) + \sqrt{2} \sqrt{\tilde \chi} \tilde \phi(t,0) = \tilde L y.
\end{equation}
Equations (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
\begin{equation} \label{eq:3.14}
	\dot {y} = \mathbb J_{2k} \tilde L^T \tilde f(t), \quad \tilde f(t) + \int_0^t \tilde \chi\cdot \tilde f(s)\ ds = \tilde L y.
\end{equation}
Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a})-(\ref{eq:3.12.b}).

We point out that the transformation that connects (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
\begin{equation}
	\mathcal A = \begin{pmatrix}
		A& 0 \\
		0& A
	\end{pmatrix} : \mathbb R^{2n} \times \mathcal H^{2n} \to \mathbb R^{2k} \times \mathcal H^{2k}.
\end{equation}
This is a symplectic transformation, since $\mathcal A^T \mathcal J_{2n} \mathcal A = \mathcal J_{2k}$. Furthermore, the dissipation of energy in the reduced system only depends on the reduced susceptibility. Thus, the choice of $A$ should be independent of the hidden strings $(\phi, \theta)$. In other words, if the reduced space is chosen to be a symplectic subspace, then the actions of model reduction and Hamiltonian extension commute. We summarize the algorithm for model reduction of dissipative Hamiltonian systems in Algorithm \ref{alg:3.1}.

\begin{algorithm}
\caption{The Reduced Dissipative Hamiltonian Method (RDH)} \label{alg:3.1}
\begin{enumerate}
	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
	\item Construct an ortho-symplectic basis $A$.
	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.c})
\end{enumerate}
\end{algorithm}

@@

{\edit The following theorem provides an indication for the boundedness of the reduced system.
\begin{theorem}
Consider the TDD system (\ref{eq:2.4}) together with the initial condition (\ref{eq:2.10.1}). If in the absence of dissipation $(\chi = 0)$, a bounded open neighborhood $U\subset \mathbb R^{2n}$ of $z_0$ exists such that $H(z)<c$, for some constant value $c > H(z_0)$, then the RDH method yields a bounded reduced system.
\end{theorem}
\begin{proof} For a Hamiltonian of the form (\ref{eq:3.7}), $H(z)\geq 0$ for all $z\in \mathbb R^{2n}$. It is shown in \cite{Figotin:2005,Figotin:2006jy} that for a symmetric and positive definite $\chi$, $\frac d {dt} H(z(t)) \leq 0$. Therefore, for all $0\leq t <\infty$, $H(z) < c$ which implies $z(t) \in U$. Now suppose that the symplectic basis $A$ is used to construct the RDH reduced system (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) such that $y_0 = A^+z_0$. It is easily checked that $V = \text{Range}(A)\cap U$ is a bounded open neighborhood of $z_0$ and that $H(z) = H(Ay) < c$ for all $z \in V$. Since $\tilde H(y) = H(Ay)$ is a Hamiltonian function for the dissipative Hamiltonian system (\ref{eq:3.14}) and that $\tilde \chi$ is symmetric and positive definite, then $\frac d {dt} \tilde H(y) \leq0$. This implies that $\tilde H(y)<c$ and $y(t)\in U$ for all $0\leq t <\infty$. Also it is easy to check that the boundedness of the solution of the original TDD system implies boundedness of the extended system in $\mathbb R^{2n}\times \mathcal H^{2n}$ with respect to the norm $\| (x,f) \|_{\text{ex}} = \|x\|_2 + \| f \|_{\mathcal H ^{2n}}$.
\end{proof}

Note that in practice, ensuring $\frac d {dt} H(z(t)) \leq 0$, requires a careful time integration of the dissipative system. The symplectic time integration of the extended system guarantees the correct evolution of the Hamiltonian.

We conclude this section by showing that the RDH method preserves the stability of the equilibrium points of the extended system. This provides a strong indication that the overall dynamics and stability of the original system system is translated into the RDH system.

\begin{definition}
Consider a dynamical system of the form $\dot z = f( z)$ and suppose that $z_e$ is an equilibrium point for the system so that $ f( z_e) = 0$. $z_e$ is called Lyapunov stable if, for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\psi_t$, if $\| \psi_0 - z_e \|_{\text{ex}} \leq \delta$, then for all $0 \leq t < \infty$, we have $\| \psi_t - z_e \|_{\text{ex}} < \epsilon$.
\end{definition}

The Dirichlet's theorem \cite{bhatia2002stability}, states that the sufficient condition for an equilibrium point to be stable is the existence of a Lyapunov function \cite{bhatia2002stability}. In the context of the Hamiltonian systems, the Hamiltonian is a candidate for the Lyapunov function.

\begin{theorem}
Consider the Hamiltonian system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) together with the reduced system (\ref{eq:3.12.a})-(\ref{eq:3.12.c}). Suppose that $(z_e,\phi_e,\theta_e)$ is an equilibrium point of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and that $z_e$ is in the range of the symplectic basis $A$. Further, suppose that $H_{\text{ex}}$ is a Lyapunov function ($\nabla^2_{(z,\phi,\theta)}H_{\text{ex}}$ is a positive definite operator). Then $(z_e,\phi_e,\theta_e)$ and $(A^+z_e,A^+\phi_e,A^+\theta_e)$ are lyapunov stable points of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), respectively.
\end{theorem}
\begin{proof} It is easily checked that $(z_e,\phi_e,\theta_e)$ is a local minimum of $H_{\text{ex}}(z,\phi,\theta)$ and a lypunov stable point for (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and also $(A^+z_e,A^+\phi_e,A^+\theta_e)$ is a local minimum of $\tilde H(y,\tilde \phi,\tilde \theta) = H_{\text{ex}}(Ay,A\tilde \phi,A\tilde \theta)$. We notice that
\begin{equation}
	\nabla^2_{(y,\tilde \phi,\tilde \theta)}\tilde H_{\text{ex}} = \mathcal A^T \nabla^2_{(z,\phi,\theta)}H_{\text{ex}} \mathcal A,
\end{equation}
is positive definite. Therefore, $\tilde H$ is a lyapunov function for (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and $(A^+z_e,A^+\phi_e,A^+\theta_e)$ is a Lyapunov stable point of  (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).
\end{proof}
}

### 2017-12-05 dissipative_hamiltonian_revision/chapters/4.numer.tex
dissipative Hmiltonian revision submitted

We construct the RDH reduced system according to the Algorithm \ref{alg:3.1}. The performance of the method is then compared to the POD and the method proposed in \cite{peng2016geometric}, referred to as the PSD.

@@

The second numerical experiment is the dissipative wave equation (\ref{eq:4.1}) in a near-zero dissipation regime. The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 10^{-5}$, for $i=1,\dots,N_{\Delta x}$. Figure \ref{fig:4.1}.(f) shows the $L^2$-error between the solution to the reduced system and the full system, for the POD, the PSD, and the RDH methods. We notice that the POD does not yield a stable reduced system as the symplectic structure is lost via model reduction. Furthermore we notice that error for the PSD and the RDH coincide as the two methods become identical as $\| \chi \|_{\infty}\to 0$.

>>>>

{\edit
We construct the RDH reduced system according to the Algorithm \ref{alg:3.1} using a basis constructed by the greedy basis selection described in Algorithm \ref{alg:3.1}. The performance of the method is then compared to the POD and the PSD proposed in \cite{peng2016geometric}. The PSD method constructs a symplectic basis using the cotangent lift method \cite{peng2016geometric}. Note the cotangent lift method can also be used to construct a basis for the RDH method. However, the greedy method and the cotangent lift yield very similar results, inline with the results in previous works \cite{Maboudi:2016}.
}

@@

The second numerical experiment is the dissipative wave equation (\ref{eq:4.1}) in a near-zero dissipation regime. The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 10^{-5}$, for $i=1,\dots,N_{\Delta x}$. Figure \ref{fig:4.1}.(f) shows the $L^2$-error between the solution to the reduced system and the full system, for the POD, the PSD, and the RDH methods. We notice that the POD does not yield a stable reduced system as the symplectic structure is lost via model reduction. Furthermore we notice that error for the PSD and the RDH coincide as the two methods become identical as $\| \chi \|_{\infty}\to 0$. {\edit Note that in this case the basis for the RDH method and PSD method are both generated using the cotangent lift method in order to show the convergence}.

### 2017-12-05 dissipative_hamiltonian_revision/paper.tex
dissipative Hmiltonian revision submitted

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{color}
\usepackage{algorithm}
\usepackage{subfig}

@@

@@

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional reduced basis methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present an approach for reduced model construction, that preserves the symplectic symmetry of dissipative Hamiltonian systems. The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of the total energy and, ultimately, in the stability of the solution. Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\let\proof\relax
\let\endproof\relax
\usepackage{amsthm}
\usepackage{color}
\usepackage{algorithm}
\usepackage{subfig}

@@

\newcommand{\edit}{\color{black}}

@@

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional reduced basis methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present an approach for reduced model construction, that preserves the symplectic symmetry of dissipative Hamiltonian systems. The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of the total energy and, ultimately, {\edit it helps conserving the} stability of the solution. Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.
